---
title: "Analysis of Teaching Methods Using ANOVA"
author: "TURRI Valentín"
date: "2025-06-03"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# Global knitr options
knitr::opts_chunk$set(echo = TRUE)

# Required libraries
library(readr)
library(knitr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(factoextra)
library(rrcov)
library(cluster)
library(FactoMineR)
library(car)
```

## *Problem Context*

An educational researcher wants to evaluate the effectiveness of three different methods for teaching mathematical problem-solving. Eighteen students with similar characteristics were randomly selected and randomly assigned to one of the three methods (6 students per group). After 4 weeks of instruction, a standardized test was administered. The obtained scores were:

```{r datos, echo=TRUE}
# Read CSV file
ruta <- "C:/Users/valen/OneDrive/Escritorio/Maestría en Ciencia de Datos/AID/TP5/datos.csv"

estudiantes_stats <- read_csv(ruta, locale = locale(encoding = "UTF-8"))

estudiantes_stats
```

## *1. Exploratory Analysis*

*A) Compute for each method: Mean, Median, and Standard Deviation*

```{r Analisis_Exploratorio, echo=TRUE}
estudiantes_stats %>%
  pivot_longer(cols = everything(), 
               names_to = "metodo", 
               values_to = "puntaje") %>%
  group_by(metodo) %>%
  summarise(
    mean = mean(puntaje, na.rm = TRUE),
    median = median(puntaje, na.rm = TRUE),
    std_dev = sd(puntaje, na.rm = TRUE)
  )
```

*B) Build a comparative boxplot for the three methods*

```{r Boxplot, echo=TRUE}
estudiantes_stats %>%
  pivot_longer(cols = everything(),
               names_to = "metodo",
               values_to = "puntaje") %>%
  ggplot(aes(x = metodo, y = puntaje, fill = metodo)) +
  geom_boxplot() + 
  labs(title = "Comparison of Teaching Methods",
       x = "Method",
       y = "Score") +
  theme_minimal() + 
  theme(legend.position = "none")
```

*C) Based on the descriptive statistics and the boxplot, formulate a preliminary hypothesis about which method might be more effective*

According to the preliminary exploratory analysis, the three methods show fairly similar effectiveness, with a 10-point difference between the highest mean score (collaborative method) and the lowest mean score.

More rigorously, the collaborative method has the highest average score, followed by the technological method, and finally the traditional method.

In all cases, the standard deviation is approximately 2 points, indicating that the scores are highly concentrated around the mean.

## *2. Assumption Verification*

ANOVA requires three fundamental assumptions, which are verified below.

### A) *Independence*

The experimental design guarantees independence of observations for the following reasons:

- **Random assignment:** Students were randomly assigned to the three teaching methods.
- **Separate groups:** Each student participated in only one method.
- **Individual assessment:** The standardized test was taken individually after instruction.

### B) *Normality*

Normality is evaluated using the Shapiro-Wilk test.

```{r Shapiro_Wilk, echo=TRUE}
datos_long <- estudiantes_stats %>%
  pivot_longer(cols = everything(), names_to = "metodo", values_to = "puntaje")

by(datos_long$puntaje, datos_long$metodo, shapiro.test)
```

Since all p-values are greater than 0.05, normality is not rejected for any group.

```{r Residuals, echo=TRUE}
modelo_anova <- aov(puntaje ~ metodo, data = datos_long)
residuos <- residuals(modelo_anova)
shapiro.test(residuos)
```

```{r QQplot, echo=TRUE}
qqnorm(residuos)
qqline(residuos, col = "red", lwd = 2)
```

The Q-Q plot shows that residuals closely follow the normal distribution.

### C) *Homoscedasticity*

```{r Levene, echo=TRUE}
leveneTest(puntaje ~ metodo, data = datos_long)
```

The p-value equals 1, so equal variances are assumed.

## *3. Analysis of Variance*

### A) Hypotheses

\[
H_0: \mu_1 = \mu_2 = \mu_3
\]

\[
H_1: \exists\, i, j \text{ such that } \mu_i \ne \mu_j
\]

### B) Why not multiple t-tests?

Multiple t-tests increase the probability of committing a Type I error. ANOVA controls this risk by performing a single global test.

### C) Interpretation of the F statistic

The F statistic compares variability between groups to variability within groups. Large F values indicate that at least one group mean differs significantly.

### D) Decision rule

```{r ANOVA, echo=TRUE}
summary(modelo_anova)
qf(0.95, df1 = 2, df2 = 15)
```

Since the observed F is greater than the critical F and the p-value is below 0.05, the null hypothesis is rejected.

## *4. Post-Hoc Comparisons*

```{r Tukey, echo=TRUE}
tukey <- TukeyHSD(modelo_anova)
tukey
plot(tukey)
```

All method pairs differ significantly.

## *5. Critical Analysis*

### Limitations

- Small sample size (18 students)
- Possible artificial data construction
- Perfectly equal standard deviations
- Extremely regular mean spacing

### Extreme outlier discussion

Robust methods (e.g., Kruskal-Wallis) may be used, or justified exclusion if measurement error is proven.
