---
title: "Principal Component Analysis of FIFA 2024 Men Player Attributes"
author: "TURRI Valentín"
date: "2025-05-10"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# Global knitr options
knitr::opts_chunk$set(echo = TRUE)

# Required libraries
library(readr)
library(knitr)
library(dplyr)
library(stringr)
library(ggplot2)
library(factoextra)
library(rrcov)
library(cluster)
```

## 1. Data

```{r data, echo=TRUE}
"The objective of this practical assignment is to apply PCA to reduce the dimensionality
of the FIFA 2024 Men dataset, identify patterns in the data, and visualize how players
are grouped according to their characteristics."

# Define path and read CSV
path <- "C:/Users/valen/OneDrive/Escritorio/Maestría en Ciencia de Datos/AID/TP3/player_stats.csv"
player_stats <- read_csv(path, locale = locale(encoding = "UTF-8"))
```

```{r seed, echo=TRUE}
"A random sample is taken from the FIFA 2024 Men database. The seed is set using the
numbers 920."

set.seed(920)
player_sample <- sample_n(player_stats, 2000)

# Display the sample
head(player_sample)

# Show the table in the HTML document:
set.seed(123)  # Change to your last 3 digits
player_sample <- sample_n(player_stats, 2000)

kable(head(player_sample))
```

# 2. Exploratory Data Analysis and Pre-Processing

## 2.1. Analyze the presence of missing or duplicated values.

```{r missing_values, echo=TRUE}
# We start by verifying the presence (or not) of missing values and/or duplicated data.

# Identify rows with at least one NA
rows_with_na <- player_sample[!complete.cases(player_sample), ]

# How many rows have NA?
cat("Number of rows with at least one NA:", nrow(rows_with_na), "\n")

# For each row with NA, indicate which columns contain NA
if (nrow(rows_with_na) > 0) {
  na_by_row <- apply(rows_with_na, 1, function(x) names(x)[is.na(x)])
  print(na_by_row)
} else {
  cat("There are no rows with missing values.\n")
}

# Duplicated rows (all columns identical)
duplicates <- duplicated(player_sample)
cat("Number of duplicated rows:", sum(duplicates), "\n")

# Check for duplicated columns
duplicated_cols <- duplicated(as.list(player_sample))  # Compare columns as lists
duplicate_column_names <- names(player_sample)[duplicated_cols]

if (length(duplicate_column_names) > 0) {
  cat("Duplicated columns found:\n")
  print(duplicate_column_names)
} else {
  cat("There are no duplicated columns.\n")
}

"- It is observed that there are 45 rows with missing values. These rows correspond to the 'Marking' column."
"- It is observed that there are no fully duplicated rows."
"- It is observed that there are no duplicated columns."
```

## 2.2. Select a subset of relevant variables to perform PCA. Justify the choice.

```{r variable_selection, echo=TRUE}
# We select numerical variables that reflect the physical, technical, and overall performance
# characteristics of outfield players. These variables include attributes such as height,
# weight, age, offensive and defensive skills, physical capabilities (speed, agility,
# strength, etc.), and technical aspects (passing, shooting, dribbling, etc.).
#
# The goal of this choice is to comprehensively capture each player's sports profile in order
# to reduce the dimensionality of the dataset without losing relevant information. In addition,
# the variable "value" is included, as it acts as an economic synthesis of the player's skills,
# incorporating an additional valuation dimension into the analysis.

# Remove $ and thousand separators, then convert to numeric
player_sample <- player_sample %>%
  mutate(value = as.numeric(gsub("[$.]", "", value)))

vars_pca <- player_sample %>%
  select(height, weight, age,
         ball_control, dribbling, slide_tackle, stand_tackle,
         aggression, reactions, att_position, interceptions, vision, composure,
         crossing, short_pass, long_pass, acceleration, stamina, strength,
         balance, sprint_speed, agility, jumping, heading, shot_power,
         finishing, long_shots, curve, fk_acc, penalties, volleys,
         value)

# Check structure
str(vars_pca)
```

## 2.3. Description of the selected variables.

```{r variable_description, echo=TRUE}
# Create a table with variable descriptions
variable_descriptions <- data.frame(
  Variable = c("height", "weight", "age", "ball_control", "dribbling",
               "slide_tackle", "stand_tackle", "aggression", "reactions",
               "att_position", "interceptions", "vision", "composure",
               "crossing", "short_pass", "long_pass", "acceleration",
               "stamina", "strength", "balance", "sprint_speed", "agility",
               "jumping", "heading", "shot_power", "finishing", "long_shots",
               "curve", "fk_acc", "penalties", "volleys", "value"),
  Description = c(
    "Player height (cm).",
    "Player weight (kg).",
    "Player age (years).",
    "Ball control when receiving.",
    "Ability to carry the ball and evade opponents.",
    "Effectiveness in sliding tackles.",
    "Effectiveness in standing tackles.",
    "Level of aggressiveness and determination.",
    "Reaction time to stimuli.",
    "Offensive positional awareness.",
    "Ability to anticipate and intercept passes.",
    "Ability to spot and assist teammates.",
    "Performance under pressure.",
    "Accuracy of crosses from the wings.",
    "Accuracy of short passes.",
    "Accuracy of long passes.",
    "Initial acceleration.",
    "Physical capacity to sustain effort.",
    "Player physical strength.",
    "Body stability while moving.",
    "Speed in sustained runs.",
    "Agility and ability to change direction.",
    "Jumping ability.",
    "Heading ability.",
    "Shot power.",
    "Finishing ability.",
    "Ability to shoot from long range.",
    "Curve applied to shots.",
    "Free-kick accuracy.",
    "Penalty accuracy.",
    "Volley accuracy.",
    "Player market value."
  ),
  stringsAsFactors = FALSE
)

# Display the table
kable(variable_descriptions, caption = "Description of selected numerical variables for PCA")
```

## 2.4. Were the variables standardized? Justify.

```{r standardization, echo=TRUE}
# Standardization of numerical variables
vars_pca_scaled <- scale(vars_pca)

# Check summary of standardized data
summary(vars_pca_scaled)

"The selected numerical variables are standardized before applying PCA. This is justified
because the variables have different scales (for example, height is measured in centimeters,
weight in kilograms, and market value in millions)."

"Without standardization, variables with larger scales would dominate the analysis,
affecting the interpretation of the principal components."
```

# 3. PCA and Interpretation

## 3.1. Perform PCA on the selected variables. How many components do you keep after dimensionality reduction?

```{r pca_analysis, echo=TRUE}
# Perform PCA on standardized data
pca_result <- prcomp(vars_pca_scaled, center = TRUE, scale. = TRUE)

# PCA summary: proportion of explained variance by each component
summary(pca_result)

# Scree plot (elbow plot)
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Explained variance by principal component") +
  theme_minimal()

# The first 10 principal components are retained because they explain more than 90% of the variance.
```

## 3.2. Interpret the loadings of the first principal components. Which player attributes are most represented in each component? Show the PCA biplot.

```{r loadings, echo=TRUE}
# Factor loadings
loadings <- pca_result$rotation[, 1:2]

# Show most influential variables for PC1 and PC2
cat("Variables with the highest influence on PC1:\n")
print(sort(abs(loadings[, 1]), decreasing = TRUE)[1:10])

cat("\nVariables with the highest influence on PC2:\n")
print(sort(abs(loadings[, 2]), decreasing = TRUE)[1:10])

# Biplot
fviz_pca_biplot(pca_result,
                geom.ind = "point",      # Points for individuals (players)
                col.ind = "steelblue",
                pointshape = 21,
                fill.ind = "lightblue",
                col.var = "tomato",      # Variable arrows color
                arrowsize = 0.8,
                repel = TRUE) +
  theme_minimal() +
  ggtitle("PCA Biplot - Components 1 and 2")

pca_result$rotation[, 1:2]

# Extract and sort loadings for PC1
loadings_pc1 <- pca_result$rotation[, 1] %>%
  sort(decreasing = TRUE) %>%
  head(10) %>%
  round(3)

# Create table
table_pc1 <- tibble::tibble(
  Variable = names(loadings_pc1),
  `Loading PC1` = loadings_pc1
)

kable(table_pc1, caption = "Main loadings of the First Component (PC1)")

# Extract and sort loadings for PC2
loadings_pc2 <- pca_result$rotation[, 2] %>%
  sort(decreasing = TRUE) %>%
  head(10) %>%
  round(3)

# Create table
table_pc2 <- tibble::tibble(
  Variable = names(loadings_pc2),
  `Loading PC2` = loadings_pc2
)

kable(table_pc2, caption = "Main loadings of the Second Component (PC2)")

"The factor loadings analysis shows that the first principal component represents offensive technical skills,
especially related to passing, shooting, and ball control. It can be interpreted as an offensive technical quality component.
On the other hand, the second principal component represents physical and defensive attributes such as strength, height,
aggressiveness, and marking ability. It can be interpreted as a physical-defensive profile component."
```

## 3.3. Identify whether there are natural groupings of players based on the selected characteristics. Interpret and conclude.

```{r natural_clusters, echo=TRUE}
# Determine the optimal number of clusters using the elbow method
fviz_nbclust(pca_result$x[, 1:5], kmeans, method = "wss") +
  ggtitle("Elbow method to determine the optimal number of clusters")

set.seed(920)  # Reproducibility

# Apply K-means on the first 5 components
kmeans_result <- kmeans(pca_result$x[, 1:5], centers = 3, nstart = 25)

# Add cluster to the original data
player_sample$cluster <- as.factor(kmeans_result$cluster)

# Biplot colored by cluster
fviz_pca_biplot(pca_result, 
                label = "none",
                habillage = player_sample$cluster,
                palette = "jco",
                addEllipses = TRUE,
                repel = TRUE) +
  ggtitle("PCA Biplot with player clusters")

"The blue cluster shows players with positive values on PC2 and moderate values on PC1. They could represent balanced players,
perhaps midfielders or box-to-box midfielders. The yellow cluster shows players with high scores on PC1, which can be interpreted
as offensive or technical players (e.g., forwards). Finally, the gray cluster shows values strongly shifted to the left on PC1.
These players likely have lower overall performance. They could be young players, low-level defenders, or simply players with
lower attribute values."
```

## 3.4. Choose a robust PCA technique and apply it to the dataset. Compare the results with the previous ones. Conclude.

```{r robust_pca, echo=TRUE, message=FALSE, warning=FALSE}
# Apply Robust PCA keeping all components
library(rrcov)
pca_robust <- PcaHubert(vars_pca_scaled, k = ncol(vars_pca_scaled), scale = FALSE, crit.pca.distances = 0.999)

# Create a prcomp-compatible object
fake_prcomp <- list(
  sdev = sqrt(pca_robust@eigenvalues),
  rotation = pca_robust@loadings,
  center = pca_robust@center,
  scale = pca_robust@scale,
  x = pca_robust@scores
)
class(fake_prcomp) <- "prcomp"

# Ensure same number of components as classical PCA
n_comp <- min(length(pca_result$sdev), length(pca_robust@eigenvalues))

# Compare explained variances
variance_comparison <- data.frame(
  Component = 1:n_comp,
  Classical = round((pca_result$sdev[1:n_comp]^2 / sum(pca_result$sdev^2)) * 100, 2),
  Robust = round((pca_robust@eigenvalues[1:n_comp] / sum(pca_robust@eigenvalues)) * 100, 2)
)

kable(variance_comparison, caption = "Comparison of explained variance")

# Compare loadings (first 2 components only)
loading_comparison <- data.frame(
  Variable = rownames(pca_result$rotation),
  PC1_Classical = round(pca_result$rotation[,1], 3),
  PC1_Robust = round(pca_robust@loadings[,1], 3),
  PC2_Classical = round(pca_result$rotation[,2], 3),
  PC2_Robust = round(pca_robust@loadings[,2], 3)
)

kable(head(loading_comparison, 10), caption = "Comparison of factor loadings")

# Adapted biplot
fviz_pca_biplot(fake_prcomp,
                geom.ind = "point",
                col.ind = "steelblue",
                col.var = "tomato",
                repel = TRUE) +
  ggtitle("Robust PCA Biplot") +
  theme_minimal()

# Clustering with the same number of components
set.seed(920)
kmeans_robust <- kmeans(pca_robust@scores[,1:n_comp], centers = 3, nstart = 25)
player_sample$cluster_robust <- as.factor(kmeans_robust$cluster)

# Comparative visualization
gridExtra::grid.arrange(
  fviz_pca_biplot(pca_result,
                  habillage = player_sample$cluster,
                  palette = "jco",
                  title = "Classical PCA"),
  fviz_pca_biplot(fake_prcomp,
                  habillage = player_sample$cluster_robust,
                  palette = "jco",
                  title = "Robust PCA"),
  ncol = 2
)

# Classical PCA loadings (first 2 components)
classical_loadings <- pca_result$rotation[, 1:2] %>%
  as.data.frame() %>%
  rename(PC1_Classical = PC1, PC2_Classical = PC2) %>%
  mutate(Variable = rownames(.)) %>%
  select(Variable, everything())

# Robust PCA loadings (first 2 components)
robust_loadings <- pca_robust@loadings[, 1:2] %>%
  as.data.frame() %>%
  rename(PC1_Robust = PC1, PC2_Robust = PC2) %>%
  mutate(Variable = rownames(.)) %>%
  select(Variable, everything())

# Combine both tables
loadings_comparison_full <- classical_loadings %>%
  inner_join(robust_loadings, by = "Variable") %>%
  mutate(
    Difference_PC1 = abs(PC1_Classical) - abs(PC1_Robust),
    Difference_PC2 = abs(PC2_Classical) - abs(PC2_Robust)
  ) %>%
  arrange(desc(abs(PC1_Classical)))

# Display comparative table
kable(loadings_comparison_full, digits = 3, caption = "Comparison of factor loadings between classical and robust PCA")

# Main differences analysis
cat("Variables with the largest change in PC1 (Classical vs Robust):\n")
top_changes_pc1 <- loadings_comparison_full %>%
  arrange(desc(abs(Difference_PC1))) %>%
  head(5)
kable(top_changes_pc1)

cat("\nVariables with the largest change in PC2 (Classical vs Robust):\n")
top_changes_pc2 <- loadings_comparison_full %>%
  arrange(desc(abs(Difference_PC2))) %>%
  head(5)
kable(top_changes_pc2)

# Interpretation
"Key Insights:
1. Technical variables such as 'short_pass' and 'vision' gain importance in robust PCA for PC1
2. Physical attributes ('strength', 'jumping') show lower relative weight in the robust method
3. The variable 'value' keeps similar importance in both methods
4. Greater stability is observed in defensive variables ('interceptions', 'stand_tackle')
5. Robust PCA reduces the influence of outliers in extreme attributes ('sprint_speed', 'shot_power')"
```
