---
title: "Correspondence and Network Analysis of Obesity-Related Data"
author: "TURRI Valentín"
date: "2025-05-10"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# Global knitr options
knitr::opts_chunk$set(echo = TRUE)

# Required libraries
library(readr)
library(knitr)
library(dplyr)
library(stringr)
library(ggplot2)
library(factoextra)
library(rrcov)
library(cluster)
library(FactoMineR)
```

## 1. Data

We start by reading the data. The goal is to perform a correspondence analysis
to explore relationships between weight and age.

```{r data, echo=TRUE}
# Read CSV file
path <- "C:/Users/valen/OneDrive/Escritorio/Maestría en Ciencia de Datos/AID/TP4/obesidad.csv"
obesity_stats <- read_csv(path, locale = locale(encoding = "UTF-8"))

# Random sample of 500 cases
set.seed(920)
obesity_sample <- obesity_stats %>% sample_n(500)

str(obesity_sample)
```

```{r categorization, echo=TRUE}
obesity_sample <- obesity_sample %>%
  mutate(
    Weight_cat = cut(
      Weight,
      breaks = c(0, 50, 70, 90, 200),
      labels = c("Low", "Normal", "High", "Very High"),
      right = FALSE
    ),
    Age_cat = cut(
      Age,
      breaks = c(0, 18, 30, 50, 100),
      labels = c("Young", "Young Adult", "Adult", "Older"),
      right = FALSE
    )
  )
```

A contingency table is created.

```{r contingency, echo=TRUE}
# Create contingency table
weight_age_table <- table(obesity_sample$Weight_cat, obesity_sample$Age_cat)

# Display table
kable(weight_age_table, caption = "Contingency table: Categorized Weight vs. Categorized Age")

# Relative table (row proportions)
relative_table <- prop.table(weight_age_table, margin = 1)
relative_table_df <- as.data.frame(relative_table)
colnames(relative_table_df) <- c("Weight_cat", "Age_cat", "Relative_frequency")

# Bar chart
ggplot(relative_table_df, aes(x = Weight_cat, y = Relative_frequency, fill = Age_cat)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Age distribution by weight category",
    x = "Weight category",
    y = "Relative frequency"
  ) +
  theme_minimal()
```

A Chi-square independence test is performed to assess whether there is independence
between the selected variables.

```{r variable_selection, echo=TRUE}
# Remove columns with all zeros (if any)
weight_age_table_filtered <- weight_age_table[, colSums(weight_age_table) > 0]

# Chi-square test (suppress warnings to avoid knit warnings, if desired)
chi_test <- suppressWarnings(chisq.test(weight_age_table_filtered))

# Show p-value in scientific notation
cat(paste0("**p-value (Chi-square):** ", format(chi_test$p.value, scientific = TRUE), "<br>"))
```

The p-value is very small, so the data suggest some degree of dependence.

```{r correspondence, echo=TRUE}
# Simple Correspondence Analysis (CA)
ca_result <- CA(weight_age_table_filtered, graph = FALSE)

# Biplot
fviz_ca_biplot(ca_result, repel = TRUE) +
  ggtitle("Correspondence Analysis Biplot - Weight vs Age")
```

**Interpretation (CA):**

- "Young" appears clearly separated in the upper-right quadrant, associated with "Low" and "Normal" weight categories, indicating that younger individuals tend to have lower weights.
- "Young Adult" and "Very High" are positioned close to each other, suggesting that individuals in that age group more frequently present very high weights.
- "Adult" appears relatively close to the "High" category, indicating some association between these two.
- "Normal" and "Low" are fairly close, which may indicate there is no strong age differentiation among individuals in those weight categories.
- Dimension 1 mainly separates weight levels (from "Very High" to "Low"), while Dimension 2 captures smaller differences, likely related to age extremes such as "Adult" and "Young".

```{r select_columns_for_mca, include=TRUE, message=FALSE, warning=FALSE}
# Select only categorical columns for MCA
mca_vars <- obesity_sample %>%
  select(SMOKE, Gender, family_history_with_overweight, MTRANS)

# Display the first rows
head(mca_vars)
```

```{r mca, include=TRUE, message=FALSE, warning=FALSE}
# Apply MCA
mca_result <- MCA(mca_vars, graph = FALSE)
```

```{r mca_visualization, include=TRUE, message=FALSE, warning=FALSE}
# Visualize MCA biplot
fviz_mca_biplot(
  mca_result,
  repel = TRUE,
  ggtheme = theme_minimal()
) +
  ggtitle("Multiple Correspondence Analysis Biplot")
```

```{r contributions, include=TRUE, message=FALSE, warning=FALSE}
# Variable contributions to Dimension 1
cat("Contributions to Dimension 1:\n")
print(fviz_contrib(mca_result, choice = "var", axes = 1, top = 10))

# Variable contributions to Dimension 2
cat("Contributions to Dimension 2:\n")
print(fviz_contrib(mca_result, choice = "var", axes = 2, top = 10))
```

### Interpretation (MCA)

**1. Multiple Correspondence Analysis Biplot**  
The biplot shows the projection of individuals (blue) and variable categories (red) onto the first two MCA dimensions.

**2. Contribution to Dimension 1**  
The variables contributing the most to Dimension 1 are mainly the mode of transportation and gender. The category `MTRANS_Bike` stands out (around 35% contribution), indicating a strong differentiation from the rest. It is followed by `MTRANS_Automobile` (~18%), `Gender_Male` (~11%), `Gender_Female` (~9.5%), and `MTRANS_Motorbike` (~8%). This suggests Dimension 1 is strongly influenced by transportation preferences (with "Bike" as a distinctive behavior) and also by gender.

**3. Contribution to Dimension 2**  
For Dimension 2, the most influential categories are `MTRANS_Walking` (~30%), followed by `Gender_Male` (~18%), `Gender_Female` (~17%), and `MTRANS_Bike` (~16%). This dimension seems to reflect a combination of gender and preference for more accessible/active transportation (e.g., walking). The strong loading of `MTRANS_Walking` suggests this dimension could be interpreted as representing preference or access to less motorized transportation, differentiating certain gender and behavior profiles.

